---
title: "HawkEars_species_verification"
author: "Erin Tattersall"
date: "2026-04-07"
output: html_document
---

*Note: I created this as a markdown file because I thought I would have to run Python code chunks to run HawkEars. That wasn't actually necessary, since HawkEars is run with a Command Line Interface (so can be run wrapped in the system2 function). Running this in an R file would save all of the working directory headaches!

Environment set up

```{r setup, echo = FALSE}

# A list of the required packages (not all used in this script - copied from Chris's scripts)
list.of.packages <- c("wildrtrax",
                      "tidyverse",
                      "kableExtra",
                      "webshot2",
                      "chromote",
                      "ggplot2",
                      "cowplot",
                      "RColorBrewer",
                      "here", # helps find project files (and set root directories)
                      "withr", # to temporarily change directories
                      "reticulate") #enables coding in python

# A check to see which ones I have and which are missing
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

# Code which tells R to install the missing packages
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

getwd()
here()
# Set root directory for entire document
knitr::opts_knit$set(
  root.dir = here::here()
)
getwd()

## By default, running chunks return results but silence warnings and messages. Also set the working directory to the git root directory (NOT the R/ folder containing the .Rmd)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

Load ARU tag data from WildTrax

```{r WildTrax}
## Authenticate into WildTrax. Access local script for WT_USERNAME and WT_PASSWORD (wildtrax_login.R - not shared on GitHub)
source(here("wildtrax_login.R")) ## This will set the environment variables WTUSERNAME and WTPASSWORD (adding here() ensures source searches the root directory)
wt_auth()

## Get data from relevant WildTrax projects
aru_projects <- wt_get_projects("ARU") %>%
  filter(organization_name == "CWS-NOR") # filter for CWS North projects
glimpse(aru_projects)

## There are 4 Edehzhie projects (project_id): 2016 (46), June 2019 (172), Total 2019 (2478), and 2021 (1052) -- only want the 2021 data (paired with cameras)
## There are 3 Gameti projects (project_id): June 2023 (3212), June 2024 (3213), and June - Sep 2024 (3007) -- all part of the same overall project, no overlapping recordings (just overlapping June periods)
## There are 2 Thaidene Nene projects: 1254 (Thaidene Nene PA) and 1232 (Boreal Monitoring Strategy)
## Other projects (project_id): 2333 (Fort Smith), 2373 (Norman Wells), 1696 (Sambaa K'e)

## Filter to my target projects only, using project IDs:
aru_projects <- aru_projects %>% filter(project_id == "1052" |
                                          project_id == "3212" |
                                          project_id == "3213" |
                                          project_id == "3007" |
                                          project_id == "1254" |
                                          project_id == "1232" |
                                          project_id == "2333" |
                                          project_id == "2373" |
                                          project_id == "1696")



## Get tag reports from all projects - that is, all species tags recorded for each project
aru_tags <- wt_download_report(project_id = aru_projects$project_id,
                               sensor_id = "ARU",
                               report = "tag") # get tag reports

glimpse(aru_tags) # List of data frames

```

Create a single dataframe with tags from all projects to extract  game bird clips from full NWT data set
```{r}
aru_tags_nt <- bind_rows(aru_tags)

## Create a list of target gamebird species codes
tar_spp <- c("ROPT", "RUGR", "SPGR", "STGR", "WIPT")

## Isolate NWT gamebird tags
nt_gb <- aru_tags_nt %>% 
  filter(species_code %in% tar_spp)

# total clips per gamebird species
table(nt_gb$species_code)

## Download target species audio files, including their spectrograms
nt_tar_rec <- wt_download_media(
  input = nt_gb,
  output = "data/wildtrax_download_aru/audio/NWT_test",
  type = "tag_clip_audio")


```

### Run HawkEars (using python) to ID clips

See HawkEars documentation (https://github.com/jhuus/HawkEars) for full installation and implementation instructions.

HawkEars is coded in Python and run via command line. As such, commands need to either be executed using reticulate functions or system2 functions

First, set up a virtual environment in the Python folder using the reticulate package.


```{r}

# Activate the correct venv
use_virtualenv("C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/Python", required = TRUE)

# Check that Python environment is configured properly
py_config() 

```

Now, install HawkEars. As per HawkEars documentation (https://github.com/jhuus/HawkEars), torch, torchaudio, and torchvision also have to be re-installed
```{r}
# Install HawkEars
py_install("HawkEars", pip = TRUE)

# Uninstall torch, torchaudio, and torchvision (no uninstall pip function in R, so call python using system2 and a python executable)
system2(reticulate::py_exe(), c("-m", "pip", "uninstall", "torch", "torchaudio", "torchvision"))

## Reinstall correct versions of these packages

py_install(
  packages = c(
    "torch==2.8.0",
    "torchvision==0.23.0",
    "torchaudio==2.8.0"
  ),
  pip = TRUE,
  pip_options = "--index-url https://download.pytorch.org/whl/cu126"
)

## Temporarily set working directory to data/download to initialize HawkEars in a git ignored data folder
with_dir("data/HawkEars_download", system2("hawkears", "init"))

```

Now HawkEars can now be run using command line prompts

```{r, echo = FALSE}

## Check HawkEars analysis options
system2("hawkears", #use hawkears package
        c("analyze", "--help")) # call analyze script and --help function
```

Full HawkEars species verification test on all NWTBMP WildTrax gamebird clips

```{r}
# create an output directory
dir.create("data/wildtrax_download_aru/audio/NWT_gamebirds/he_output")

# Run HawkEars on all WT NWT game bird recordings
with_dir("data/HawkEars_download", ## Set working directory to run analyses where scripts are saved
         system2( # run command line prompt
  command = "hawkears", # run HawkEars python package
  c("analyze", # run analyze script
    "-i", "C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/data/wildtrax_download_aru/audio/NWT_gamebirds", # set input folder to recordings
    "-o", "C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/data/wildtrax_download_aru/audio/NWT_gamebirds/he_output", # set output folder
    "--recurse", # process sub-directories (none here, but important later)
    "-r", "csv", # specify output as csv
    "--region", "CA-NT" # specifies the eBird region code
    ))) 


```

Check the output for all NWT clips
```{r, echo = FALSE}

he_nt <- read.csv("data/wildtrax_download_aru/audio/NWT_gamebirds/he_output/scores.csv")

## should be 905 unique recordings
length(unique(he_nt$recording)) #572 - so HawkEars doesn't detect a bird in all of the recordings

## filter for target species only
he_gb <- he_nt %>% 
  filter(name %in% tar_spp) ## 390 gamebird calls ID'd by HE
length(unique(he_gb$recording)) # 358 - HE is detecting more than one species per clip, even though each clip represents a tag for a single species (other spp may still be calling at the same time, but multiple game birds calling is unlikely given these clips are so short)



## rename he_gb column names for clarity
colnames(he_gb) <- c("clip_name", "he_spp_ID", "start_time", "end_time", "conf_score")

## What is the total range of confidence scores?
summary(he_gb$conf_score)


table(he_gb$he_spp_ID)

# Combine HawkEars score range per species
he_scores_spp <- he_gb %>% group_by(he_spp_ID) %>% summarise(
  min_score = min(conf_score),
  mean_score = mean(conf_score),
  max_score = max(conf_score)
)


```
HawkEars confidence scores a quite high for all game bird species. How do they compare to the WildTrax tags?

A field key is needed to match HawkEars IDs to the manual tags in nt_gb. When the WT recordings were downloaded, they were named based on these fields in nt_gb:
(organization)_(location)_(recording_date_time)__(species_code)__(individual_order)__(detection_time)

We also want to select only relevant column in nt_gb
```{r, echo = FALSE}
## Add clip name to nt_gb
nt_gb$clip_name <- with(nt_gb, paste0(
  organization, "_",
  location, "_",
  format(as.POSIXct(recording_date_time), "%Y%m%d_%H%M%S"),
  "__",
  species_code,
  "__",
  individual_order,
  "__",
  gsub("\\.", "_", detection_time)
))

## Select relevant columns (move clip_name to first column)
gb2 <- nt_gb %>% 
  select(clip_name, organization, location, recording_date_time, species_code, individual_order, detection_time, vocalization, recording_id, tag_id)

length(unique(gb2$clip_name))

```
Next, we combine the WildTrax tags to HawkEars gamebird output. We can measure performance by marking true and false positives and false negatives.
```{r}

## Combine WT tags with HawkEars gamebird IDs to create a 'positives' df. Left join keeps all observations in he_gb
gb_pos <- left_join(he_gb, gb2, by = "clip_name")

## Create a observed column, where observed = 1 if the HE ID matches the species code, 0 if not
gb_pos$observed <- ifelse(gb_pos$he_spp_ID == gb_pos$species_code,
                                       yes = 1,
                                       no = 0)
table(gb_pos$observed)

## Re-order gb_pos by he_spp_ID and observed
gb_pos <- gb_pos %>%  arrange(he_spp_ID, observed)

## True and false positives per species
spp_pos <- gb_pos %>% 
  group_by(he_spp_ID) %>% ## need to group by HawkEars species, not WT
  summarise(tp = sum(observed),
            fp = sum(observed == 0))
```
Summarise results in a table. Summary stats:

HawkEars confidence score: certainty of identification, NOT probability of true detection. Lower confidence scores on true positives indicate that the algorithm is right even when it's not certain. Think Dunning-Kruger effect - high confidence often correlated with low knowledge!
Precision: The proportion of times HawkEars ID'd a species and was right ('accuracy')
Recall: The proportion of times HawkEars detected a bird and correctly ID'd it ('sensitivity')
F1: estimates how well HawkEars IDs calls relative to a human observer
```{r}

## Create a species table with total number of clips, total number of clips with HawkEars IDs, and confidence scores
# total NWT gamebird clips
gb_tbl <- as.data.frame(table(nt_gb$species_code))
## add total HawkEars IDs
gb_tbl <- left_join(gb_tbl, as.data.frame(table(he_gb$he_spp_ID)), by = "Var1")
## add HawkEars confidence scores
gb_tbl <- cbind.data.frame(gb_tbl, he_scores_spp)

## Rename columns and remove 2nd species column
colnames(gb_tbl) <- c("species_code", "total_wt_clips", "total_HawkEars_IDs", "species_code2", "HEscore_min", "HEscore_mean", "HEscore_max")
gb_tbl <- gb_tbl[ , c(1:3, 5:7)]

## Add true and false positives (first need to rename he_spp_ID)
colnames(spp_pos) <- c("species_code", "tp", "fp")
gb_tbl <- left_join(gb_tbl, spp_pos, by = "species_code") 

# Add false negatives (total_clips - true positives)
gb_tbl$fn <- gb_tbl$total_wt_clips - gb_tbl$tp


## Add precision, recall, and F1 scores
gb_tbl$precision <- gb_tbl$tp/(gb_tbl$tp + gb_tbl$fp)
gb_tbl$recall <- gb_tbl$tp/(gb_tbl$tp + gb_tbl$fn)
gb_tbl$F1 <- 2 * ((gb_tbl$precision * gb_tbl$recall)/(gb_tbl$precision + gb_tbl$recall))

gb_tbl

## Save table, but summarize as total clips, mean HE score, precision, recall, and F1. Rename for publishing
gb_tbl2 <- gb_tbl %>% 
  select("Species Code" = species_code, "No. recordings" = total_wt_clips, "Mean HawkEars confidence" = HEscore_mean, "Precision" = precision, "Recall" = recall, "F1 score" = F1)  

gb_tbl2
## save as csv in figures for formatting into a table
write.csv(gb_tbl2, "C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/figures/HawkEars_performance_WT_gamebirds_20260409.csv")
```


Plot the detection rate against HawkEars confidence scores (using code from Tseng et al. 2025)
```{r}
coul <- brewer.pal(5, "Set1")

g <- ggplot(gb_pos, aes(x = conf_score,
                        y = observed,
                        group = he_spp_ID, # use the HawkEars spp code
                        colour = he_spp_ID)) +
  geom_point(size = 5, 
             alpha = 0.1) +
  geom_line(stat = "smooth",
            method = "glm", 
            se = FALSE, 
            method.args = list(family = binomial),
            linewidth = 1.5,
            alpha = 0.7) +
  scale_colour_manual(values = coul) +
  scale_x_continuous(limits = c(0.7, 1), expand = c(0, 0), breaks = seq(0.7, 1, by = 0.1)) +
  scale_y_continuous(limits = c(0, 1)) + 
  theme_bw() +
  labs(x = "HawkEars confidence", 
       y = "True positive rate",
       colour = "Species") +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.position = "bottom",
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
        plot.margin = margin(1, 1, 1, 1, "cm")) +
  guides(colour = guide_legend(ncol = 4)) 

g

## Save plot
ggsave(plot = g,
       filename = "C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/figures/WT_recs_HE_gamebird_calibration_curves_20260409.jpeg",
       width = 24,
       height = 19,
       units = "cm",
       dpi = 300)
```

Conclusions:
- For ROPT, STGR, and WIPT, HawkEars seems to be conservative in its IDs (higher false negative rate = lower recall). F1 scores suggest that HawkEars identifies these species about 53%, 47% or 45% (respectively) as often as a human observer (i.e., HawkEars is half as good as a human).
- HawkEars is more sensitive to RUGR and SPGR, but less accurate (more FPs)
- Only willow ptarmigan (WIPT, clips = 730) met the recommended number of clips to estimate a confidence threshold (n = 180, Tseng et al. 2025). False positives for willow ptarmigan were very low (fp = 1, which had a score of 0.94, explaining the slight dip in positive rate at high confidence). HawkEars has a low WIPT detection probability, but we can be fairly certain that any WIPT ID is accurate.
- Since ROPT, STGR, and WIPT have 0 or 1 false positives, we can't really calculate a confidence threshold for those species. We may be able to for RUGR and SPGR, but only if we are comfortable with precision < 0.75. Based on the graph, we would need to set the confidence threshold at about 0.93 for RUGR, and 0.97 for SPGR to achieve a 75% true positive rate.
- Given that our sample of clips per species is lower than recommended for 4 species, we cannot be certain that these results indicate reliable confidence thresholds.


