---
title: "HawkEars_species_verification"
author: "Erin Tattersall"
date: "2026-04-07"
output: html_document
---

Environment set up

```{r setup}

# A list of the required packages (not all used in this script - copied from Chris's scripts)
list.of.packages <- c("wildrtrax",
                      "tidyverse",
                      "ggplot2",
                      "cowplot",
                      "here", # helps find project files (and set root directories)
                      "withr", # to temporarily change directories
                      "reticulate") #enables coding in python

# A check to see which ones I have and which are missing
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

# Code which tells R to install the missing packages
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)



# Set root directory for entire document
knitr::opts_knit$set(
  root.dir = here::here()
)

## By default, running chunks return results but silence warnings and messages. Also set the working directory to the git root directory (NOT the R/ folder containing the .Rmd)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

Load ARU tag data from WildTrax

```{r WildTrax}

## Authenticate into WildTrax. Access local script for WT_USERNAME and WT_PASSWORD (wildtrax_login.R - not shared on GitHub)
source("wildtrax_login.R") ## This will set the environment variables WTUSERNAME and WTPASSWORD
wt_auth()

## Get data from relevant WildTrax projects
aru_projects <- wt_get_projects("ARU") %>%
  filter(organization_name == "CWS-NOR") # filter for CWS North projects
glimpse(aru_projects)

## There are 4 Edehzhie projects (project_id): 2016 (46), June 2019 (172), Total 2019 (2478), and 2021 (1052) -- only want the 2021 data (paired with cameras)
## There are 3 Gameti projects (project_id): June 2023 (3212), June 2024 (3213), and June - Sep 2024 (3007) -- all part of the same overall project, no overlapping recordings (just overlapping June periods)
## There are 2 Thaidene Nene projects: 1254 (Thaidene Nene PA) and 1232 (Boreal Monitoring Strategy)
## Other projects (project_id): 2333 (Fort Smith), 2373 (Norman Wells), 1696 (Sambaa K'e)

## Filter to my target projects only, using project IDs:
aru_projects <- aru_projects %>% filter(project_id == "1052" |
                                          project_id == "3212" |
                                          project_id == "3213" |
                                          project_id == "3007" |
                                          project_id == "1254" |
                                          project_id == "1232" |
                                          project_id == "2333" |
                                          project_id == "2373" |
                                          project_id == "1696")



## Get tag reports from all projects - that is, all species tags recorded for each project
aru_tags <- wt_download_report(project_id = aru_projects$project_id,
                               sensor_id = "ARU",
                               report = "tag") # get tag reports

glimpse(aru_tags) # List of data frames

```
Test: Isolate tags for target species in a single study area to test downloading audio from WildTrax

```{r}
## create target species list (using species codes)
tar_spp <- c("ROPT","RUGR","SPGR", "STGR", "WIPT")

## Get raw report from Edehzhie only (4th in list)
ede_tags <- aru_tags[[2]]
glimpse(ede_tags)

## how many unique species codes?
length(unique(ede_tags$species_code)) # 140
```

```{r}
## Isolate target spp tags in Edehzhie
ede_tar <- ede_tags %>% 
  filter(species_code %in% tar_spp)


## Download target species audio files, including their spectrograms
ede_tar_rec <- wt_download_media(
  input = ede_tar,
  output = "data/wildtrax_download_aru/audio/ede_gamebirds",
  type = "tag_clip_audio")

```
### Run HawkEars (using python) to ID clips

See HawkEars documentation (https://github.com/jhuus/HawkEars) for full installation and implementation instructions.

HawkEars is coded in Python and run via command line. As such, commands need to either be executed using reticulate functions or system functions

First, set up a virtual environment in the Python folder using the reticulate package.


```{r}
# Create venv
virtualenv_create("C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/Python")

# Activate the correct venv
use_virtualenv("C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/Python", required = TRUE)

# Check that Python environment is configured properly
py_config() 

```
Now, install HawkEars. As per HawkEars documentation (https://github.com/jhuus/HawkEars), torch, torchaudio, and torchvision also have to be re-installed
```{r}
# Install HawkEars
py_install("HawkEars", pip = TRUE)

# Uninstall torch, torchaudio, and torchvision (no uninstall pip function in R, so call python using system2 and a python executable)
system2(reticulate::py_exe(), c("-m", "pip", "uninstall", "torch", "torchaudio", "torchvision"))

## Reinstall correct versions of these packages

py_install(
  packages = c(
    "torch==2.8.0",
    "torchvision==0.23.0",
    "torchaudio==2.8.0"
  ),
  pip = TRUE,
  pip_options = "--index-url https://download.pytorch.org/whl/cu126"
)

## Temporarily set working directory to data/download to initialize HawkEars in a git ignored data folder
with_dir("data/HawkEars_download", system2("hawkears", "init"))


```

Now HawkEars can now be run using command line prompts

```{r}

## Check HawkEars analysis options
system2("hawkears", #use hawkears package
        c("analyze", "--help")) # call analyze script and --help function

# Test HE with recording provided
with_dir("data/HawkEars_download", ## Set working directory to run analyses where scripts are saved
         system2( # run command line prompt
  command = "hawkears", # run HawkEars python package
  c("analyze", # run analyze script
    "-i", "recordings", # set input folder to recordings
    "-rtype", "csv" # specify output as csv
    ))) 

# Check output
list.files("data/HawkEars_download/recordings")

HE_COYE <- read.csv("data/HawkEars_download/recordings/scores.csv")

glimpse(HE_COYE)

```

Now we are ready to test HawkEars on the Edéhzhíe recordings downloaded earlier!
```{r}
## check the path to the recordings - will need to use full filepath, since the root directory will be in HawkEars_download
list.files("data/wildtrax_download_aru/audio/ede_gamebirds") # audio files are both .flac and .mp3

# create an output directory
dir.create("data/wildtrax_download_aru/audio/ede_gamebirds/he_output")

# Run HawkEars on Edehzhie game bird recordings
with_dir("data/HawkEars_download", ## Set working directory to run analyses where scripts are saved
         system2( # run command line prompt
  command = "hawkears", # run HawkEars python package
  c("analyze", # run analyze script
    "-i", "C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/data/wildtrax_download_aru/audio/ede_gamebirds", # set input folder to recordings
    "-o", "C:/Users/tatterer.stu/Desktop/nwtbm_phd_gamebirds/data/wildtrax_download_aru/audio/ede_gamebirds/he_output", # set output folder
    "--recurse", # process sub-directories (none here, but important later)
    "-r", "csv", # specify output as csv
    "--region", "CA-NT" # specifies the eBird region code
    ))) 


```
Check the output
```{r, echo = FALSE}
# Check output
list.files("data/wildtrax_download_aru/audio/ede_gamebirds/he_output")

he_ede <- read.csv("data/wildtrax_download_aru/audio/ede_gamebirds/he_output/scores.csv")

glimpse(he_ede)
## should be 140 unique recordings
length(unique(he_ede$recording)) #124 - HawkEars doesn't detect a bird in all of the recordings

## filter for target species only
he_ede <- he_ede %>% 
  filter(name %in% tar_spp) ## 111 species
length(unique(he_ede$recording)) #104 - HE is detecting more than one species per clip, even though each clip represents a tag for a single species (other spp may still be calling at the same time, though)

```
A field key is needed to match HawkEars IDs to the manual tags in ede_tar. When the WT recordings were downloaded, they were named based on these fields in ede_tar:
(organization)_(location)_(recording_date_time)__(species_code)__(individual_order)__(detection_time)

We also want to select only relevant column in ede_tar
```{r}
## Add clip name to ede_tar
ede_tar$clip_name <- with(ede_tar, paste0(
  organization, "_",
  location, "_",
  format(as.POSIXct(recording_date_time), "%Y%m%d_%H%M%S"),
  "__",
  species_code,
  "__",
  individual_order,
  "__",
  gsub("\\.", "_", detection_time)
))

## Select relevant columns (move clip_name to first column)
ede_2 <- ede_tar %>% 
  select(clip_name, organization, location, recording_date_time, species_code, individual_order, detection_time, vocalization, recording_id, tag_id)

```
Next, we combine the HawkEars output to the WildTrax tags to see how they match

